---
title: "candymini"
author: "Colin Mach"
format: html
---

```{r}
candy_file <- "candy-data.csv"
candy = read.csv(candy_file, row.names = 1)
head(candy)
```

# Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```

There are 85 different candy types

# Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```
There are 38 fruity candy types in the dataset.

```{r}
candy["Twix",]$winpercent
```
# Q3. What is your favorite candy in the dataset and what is it's winpercent value?

```{r}
candy["Sour Patch Kids",]$winpercent
```
Favorite candy is sour patch kids and its winpercent is 59.8%

# Q4 What is the winpercent value for "Kit Kat"?

```{r}
candy["Kit Kat",]$winpercent
```

# Q5. What is the winpercent value for "Tootsie Roll Snack Bars"?
```{r}
candy["Tootsie Roll Snack Bars",]$winpercent
```
```{r}
library("skimr")
skim(candy)
```

# Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

winpercent seems to be on a different scale to the majority of the other columns in the dataset since all the other columns are 1 or 0 while winpercent can range form 0 to 100.

# Q7. What do you think a zero and one represent for the candy$chocolate column

A zero means that the candy is not a chocolate while a one means that the candy is a chocolate candy

# Q8. Plot a histogram of winpercent values

```{r}
hist(candy$winpercent)
```

# Q9. Is the distribution of winpercent values symmetrical?

The distribution of winpercent values are not symmetrical

# Q10. Is the center of the distribution above or below 50%?

Center of the distribution is below 50%

# Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
mean(candy$winpercent[as.logical(candy$chocolate)])
mean(candy$winpercent[as.logical(candy$fruity)])
```
On average chocolate candy is higher ranked than fruit candy

# Q12. Is this difference statistically significant?

```{r}
t.test(candy$winpercent[as.logical(candy$chocolate)], candy$winpercent[as.logical(candy$fruity)])
```

The difference is statistically significant as indicated by the low p-value.

# Q13. What are the five least liked candy types in this set?

```{r}
library(dplyr)
candy %>% arrange(winpercent) %>% head(5)
```
Nik L Nip, Boston Baked Beans, Chiclets, Super Bubble, and Jawbusters are the least liked candies.

# Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
candy %>% arrange(desc(winpercent)) %>% head(5)
```
Reese's Peanut Butter cup, Reese's Miniatures, Twix, Kit Kat, and Snickers are the top 5 most liked candies.

# Q15. Make a first barplot of candy ranking based on winpercent values

```{r}
library(ggplot2)

ggplot(candy)+
  aes(winpercent, rownames(candy))+
  geom_col()
```

# Q16. use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(candy)+
  aes(winpercent, reorder(rownames(candy),winpercent))+
  geom_col()
```

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) 
```

# Q17. What is the worst ranked chocolate candy?

The worst ranked chocolate candy is sixlets

# Q18. What is the best ranked fruity candy?

The best ranked fruity candy is starburst


```{r}
library(ggrepel)
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```

# Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

Reeses Miniatures offer the most bang for your buck

# Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```

Top 5 most expensive candies are Nik L nip, Nestle Smarties, Ring pop, Hersheys Krackel, and Hersheys Milk Chocolate with Nik L Nip being the least popular out of the 5.

```{r}
library(corrplot)
cij <- cor(candy)
corrplot(cij)
```

# Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Fruity and chocolate are anticorrelated

# Q23. Similarly, what two variables are most positively correlated?

Bar and chocolate are most positively correlated.

```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
```

```{r}
plot(pca$x[,1:2])
```
```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```

```{r}
my_data <- cbind(candy, pca$x[,1:3])
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```
```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

```{r}
library(plotly)
```

```{r}
ggplotly(p)
```

```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```
# Q.24 What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

The original variables that are picked up strongly are fruity hard and pluribus which make sense since if you see the candy types that are on the right side of the PC1 vs PC2 analysis on the other graph are mostly fruity/hard/pluribus type candies .
